### 2019-10-10 submission/manuscript.tex
Final edits

Looking back on scientific journeys can be instructive to others who are
overwhelmed at the prospect of looking forward at their careers (1--5).
By no means is my scientific journey over, but since 2002 I have been on
a journey that I did not realize I was on. Now that the paper
introducing the mothur software package is ten years old and has become
the most cited paper published by \emph{Applied and Environmental
Microbiology} (7), it is worth stepping back and using the continued
development of mothur as a story that has parallels to many other
research stories.

I fondly recall preparing a poster for the 2002 meeting of research
groups supported by the NSF-supported Microbial Observatories Program. I
wanted to triumphantly show that I had sequenced more than 600 16S rRNA
gene sequences from a single 0.5-g sample of Alaskan soil. This was
greater sequencing depth than anyone else had achieved for a single
sample. As I was preparing the poster, I walked into the office of Jo
Handelsman, my postdoctoral research advisor, and laid out the outline
for the poster. She asked if I could add one of those ``curvy things'',
a rarefaction curve, to show where I was in sampling the community.
Rarefaction curves and attempts to estimate the taxonomic richness of
soil had become popular because of the impactful review by Jennifer
Hughes and her colleagues (8). Their seminal paper introduced the field
to operational taxonomic units (OTUs), rarefaction curves, and richness
estimates. I do not recall whether my poster had a rarefaction curve on
it, but Jo's question primed my career.

\textbf{\emph{Introducing DOTUR and friends.}} When Jo asked me to
generate a rarefaction curve for the poster, the request was not
trivial. How would I bin the sequences into OTUs? Hughes and her
colleagues did it manually and with fewer than 300 sequences. Although I
could possibly do that for my 600 sequences, my goal was to generate
1,000 sequences from the sample and to repeat that sampling effort for
other samples. I needed something that could be automated. Furthermore,
the software that Hughes used, EstimateS (4), required a series of
tedious data formatting steps to perform the analyses we were interested
in performing. I had found my first problem. How would I assign
sequences to OTUs and use that data to estimate the richness and
diversity of a sample? The second problem would involve comparing the
abundance of OTUs found in one sample to another sample. The solution to
the first problem, DOTUR (Distance-based OTUs and Richness), took us two
years to develop (9). DOTUR did two things: given a matrix quantifying
the genetic distance between pairs of sequences, it would cluster those
sequences into OTUs for any distance threshold to define the OTUs and
then it would use the frequency of each OTU to calculate a variety of
alpha diversity metrics. The solutions to the second problem would come
from our work to develop software including \(\int\)-LIBSHUFF (10), SONS
(Shared OTUs and Similarity) (11), and TreeClimber (12). Around the same
time, Catherine Lozupone and Rob Knight were developing their UniFrac
tools to compare communities with a phylogenetic rather than OTU-based
approach (13, 14). With these tools, the field of microbial ecology had
a quantitative toolbox for describing and comparing microbial
communities. Along the way Jo and I would demonstrate the utility of
such tools to answer questions like how many OTUs were there in that
sample of Alaskan soil and how many sequences were needed to sample each
of those OTUs (15)? Where were we in the global bacterial census (16)?
How does the word usage of \emph{Goodnight, Moon} compare to that of
\emph{Portrait of a Lady} and more importantly how is this relevant to
microbial ecology (17)? Most edifying were the more than 2,400 papers
that used DOTUR, SONS, TreeClimber, or \(\int\)-LIBSHUFF to facilitate
their own research questions (Web of Science, 10/1/2019). Had we waited
to solve all of the problems that plagued 16S rRNA gene sequencing, we
would still be waiting.

It is important to remember that we knew there were many problems with
16S rRNA gene sequencing. We knew there were biases from extractions and
amplification (18--23). We knew there were chimeras (24--27). We knew
that bacteria varied in their \emph{rrn} copy number. Generating a
distance matrix was a prerequisite to using my tools. This wasn't
trivial, but by cobbling together other tools it was possible. We would
assemble, trim and correct Sanger sequence reads using Chromas or STADEN
(28), align the sequences using ClustalW (29) or ARB (30), check for
chimeras using partial treeing or Bellerephon (27), and calculate a
pairwise distance matrix using DNADIST from the PHYLIP package (31). At
the time, we knew that we only had a loose concept of a species based on
these distances (32). We hoped that an OTU defined as a group of
sequences more than 97\% similar to each other would be a biologically
meaningful unit regardless of whether it fit our notion of a bacterial
species. At the time, I felt that the biggest problems that I could
solve were how to cluster the sequences into OTUs and how to use those
clusterings to test our hypotheses. The only tool available at the time
that automated the clustering step was FastGroup, which implemented an
approximation of the single linkage algorithm (33). The high cost of
sequencing was also an impediment to experimentation and analysis in
microbial ecology. It was rare for a study design to have experimental
replicates so that one could perform a statistical test to compare
treatment groups. For example, in our testing we frequently used a
dataset comparing Scottish soils from Alison McCaig and colleagues (34).
This dataset consisted of two experimental groups, each replicated three
times with 45 sequences per replicate. Although great focus has been
placed on the depth of sampling afforded by 454 and Illumina sequencing,
the true benefit of the modern sequencing platforms is the ability to
affordably sequence a large number of technical and biological
replicates. In spite of the many technical challenges, we had excuses
and heuristics to solve problems that served our needs. It is telling
that a recent review of ``best practices'' in generating and analyzing
16S rRNA gene sequences shows that we still have not solved many of
these issues and that we have even identified additional problems (35).

As we developed these tools, I found a unique niche in microbiology. My
undergraduate and graduate training as a biological engineer prepared me
to think about research questions from a systems perspective, to think
quantitatively, and to understand the value of using computer programs
to help solve problems. As an undergraduate student, I learned the
Pascal programming language and promptly forgot much of it. Although it
was a good language for teaching programming concepts, it did not catch
on outside of the classroom. Later, I learned MATLAB. Because it was an
expensive commercial programming environment and never caught on with
biologists, I also forgot much of it. Even if I forgot the programming
syntax of these languages, what these programming experiences taught me
was the logic and structure of programming. As a postdoc, I would use
these experiences to learn the Perl programming language to better
understand how LIBSHUFF, a tool for comparing the structure of two
communities, worked since it was written in Perl (36). After writing my
own version of LIBSHUFF, \(\int\)-LIBSHUFF, and seeing the speed of the
version written in C++ by my collaborator, Bret Larget, I converted my
Perl version of DOTUR into C++. At the time, the conversion from Perl to
C++ seemed like an academic exercise to learn a new language. My Perl
version only took a minute or so to process the final collection of
1,000 sequences and the C++ version took seconds. Was that really such a
big difference? In hindsight, as we now process datasets with millions
of sequences, the decision to learn C++ was critical. The ability to
pick up computer languages to solve problems, enabled by my prior
training in engineering, was a skill that was virtually unheard of in
microbiology. Today, researchers without the ability to program are at a
significant disadvantage (37).

\textbf{\emph{Introducing mothur.}} Shortly after DOTUR was published, I
received an email from Mitch Sogin, a scientist at the Marine Biology
Laboratory (Woods Hole, MA), who asked whether DOTUR could handle more
than a million sequences. Without answering his question, I asked where
he found a million sequences. Little did I know that his email would
represent another pivot in the development of these tools. His group
would be the first to use 454 sequencing technology to generate 16S rRNA
gene sequences (38). Although DOTUR could assign those sequences to
OTUs, at the scale of millions of sequences, it was slow and required a
significant amount of RAM. As I left my postdoc to start my independent
career across the state from Sogin's lab at the University of
Massachusetts in Amherst, my plan was to rewrite DOTUR, SONS,
\(\int\)-LIBSHUFF, and TreeClimber for the new world of massively
parallelized sequencing. The new tool would become mothur.

Milling about at a poster session at the 2007 ASM General Meeting in
Toronto, I ran into Mitch who asked what my plans were for my new lab. I
told him that I wanted to make a tool like ARB, a powerful database tool
and phylogenetics package (30), but for microbial ecology analysis. His
retort was, ``You and what army?'' Up to that point, I had written every
line of code and been answering many emails from people asking for help.
He was right, I would need an army. It would be difficult, but I needed
to learn to let go and share the development process with someone else.
My ``army'' ended up being Sarah Westcott who has worked on the mothur
project from its inception. Today, mothur is over 200,000 lines of code
and Sarah has touched or written nearly every line of it. Beyond writing
and testing mothur's code base, she has become a conduit for many who
are trying to learn the tools of microbial ecology. She patiently
answers questions via email and on the package's discussion forum
(\url{https://forum.mothur.org}). The community and I are lucky that
Sarah has stayed with the project for more than a decade. To be honest,
such dependency on a single person makes the project brittle. In
hindsight, it would have been better to have developed mothur with more
of an ``army'' or team so that there is overlap in people's
understanding of how mothur works. Although a distributed team approach
might work in a software engineering firm, it is not practical in most
academic environments where there is limited funding. There are
certainly projects that make this work, but they are rare.

@@

step. Third, as I mentioned above, there has been a plethora of methods

@@

\textbf{\emph{Challenges of making open source count.}} Anyone can post
code to GitHub with a permissive license and claim to be an open source
software developer. Far more challenging is engaging the target
community to make contributions to that code. Frankly, we have struggled
to expand the number of people that make contributions to the mothur
code base. One challenge we face is that if we looked to third parties
to contribute code to mothur, they would need to know C++. Given the
paucity of microbiologists that can program in a compiled language like
C++, expecting that community to provide contributors who can write code
in a syntax that prizes execution efficiency over developer efficiency
was not realistic. In contrast, the QIIME development team could be more
distributed because their code base was primarily written in Python,
which prizes developer efficiency over execution efficiency. QIIME is a
series of wrappers that allow users to execute other developers' code,
making the use of a scripting language like Python attractive. Their
choices resulted in many tradeoffs that have impacted ease of
installation, usability, execution speed, and flexibility. If we were
offered funding to rewrite mothur, we would likely rewrite it as an R
package that leaned heavily on the R language's C++ interface packages.
Of course, such choices are always best in hindsight. Yet, when we
started developing mothur, the ability to interface between scripting
languages like R and Python and C++ code was not as well developed as it
is today. For example, the modern version of the Rcpp package was first
released in 2009 and its popularity was not immediate (66). The
development of mothur has been a product of the environment that it was
created in. Although these decisions have largely had positive outcomes,
there have been tradeoffs that caused us to sacrifice other goals.

@@

\textbf{\emph{Failed experiments.}} If we never failed, we would not be
trying hard enough. Over the past decade we have tried a number of
experiments to improve the usability and utility of mothur. One of our
first experiments was to use mothur to generate standard vector graphic
(SVG)-formatted files of heatmaps and Venn diagrams depicting the
overlap between microbial communities. Such visuals were helpful for
exploring data; however, I quickly realized that I would never put a
mothur-generated figure into a manuscript I wrote. Such visuals require
far too much customization to be publication-quality. Although QIIME has
incorporated visualization tools through the Emperor package (67), the
challenge of users taking default values has downsides, especially when
those defaults do not follow good data visualization principles. For
example, ordinations with black backgrounds and 3-D ordinations in a 2-D
medium now litter the literature. Instead, we have encouraged users to
use R packages to visualize mothur-generated results using the minimalR
instructional materials that I have developed
(\url{http://www.riffomonas.org/minimalR/}). A second experiment was the
creation of a graphical user interface (GUI) for running mothur. Forcing
users to interact with mothur through the command line has been a
significant hurdle for many. Unfortunately, the development effort
required to create and maintain a GUI is significant and there is
limited funding for such efforts. The newest version of QIIME (starting
with version 2.0.0) has emphasized interaction with the tools through a
GUI (68) and the related QIITA project offers a web-based GUI (69). It
remains to be seen how this experiment will go. Another downside of
using a GUI is that there is a risk that reproducibility will suffer if
users do not have a mechanism to document their mouse clicks.
Documentation of commands and parameter values is explicit in users can
provide a file with a list of commands and the software returns a
logfile with all commands and output recorded. Given the heightened
focus on reproducibility in recent years, we have extended significant
effort in developing instructional materials teaching users how to
organize, document, and execute reproducible pipelines that allow a user
to go from raw sequence data to a compiled manuscript (65, 70). A final
example of a failed experiment was a collaboration with programmers
through Google Summer of Code to develop commands in mothur that ran the
random forest and SVM machine learning algorithms. Similar to the
challenges of developing attractive visuals, fitting the algorithms'
hyperparameters, testing, and deploying the resulting models require a
significant amount of customization. Furthermore, machine learning is an
active area of research where methods are still being developed and
improved. Thankfully, there are numerous R and Python packages that do a
better job of developing these models (71, 72). In each of our
``failed'' experiments, the real problems were straying from what mothur
does well and failing to grasp what we really wanted the innovation to
do.

@@

As with anything in science, methods become \text{pass\'e}. When we
first developed mothur, T-RFLP and DGGE were still commonly used. Today
it would be hard to argue that data from those methods meaningfully
advance a study relative to what one could get using 16S rRNA gene
sequence data. Looking forward, many want to claim that amplicon
sequencing is today's DGGE. They claim that researchers should instead
move on to shotgun metagenomic sequencing. It is important to note that
the two methods answer fundamentally different questions. 16S rRNA gene
sequence data describes the taxonomic composition, while metagenomic
sequence data tells a researcher about the functional potential and
genetic diversity of a community. Both tools provide important
information, but they cannot easily replace each other. Although
metagenomic data does provide highly resolved taxonomic information, the
limit of detection is at least an order of magnitude higher than that of
amplicon data. For example, we analyzed 10,000 16S rRNA sequences from
each of about 500 subjects (73). We can think of this as representing
about 1,000,000 genome equivalents (10,000 16S rRNA genes/subject x 500
subjects / 5 16S rRNA gene sequences/genome). Assuming a genome is 4
Mbp, this would represent a sequencing depth of 4 Tbp. Although such a
sequencing effort is technically possible, the cost of such an endeavor
would be considerable and unlikely to be pursued by most researchers. We
estimate that generating and sequencing the libraries at the University
of Michigan sequencing core would cost approximately \$150 per library.
The parallel 16S rRNA gene sequences data would cost approximately \$8
per library. Furthermore, analyzing such a large dataset with an
approach that captures the full genetic diversity of the community would
be financially and technically prohibitive. Going forward, sequencing
technologies will continue to evolve to capture longer and more high
quality data and there will always be a need for characterizing the
taxonomic diversity of microbial communities. With this in mind, there
will always be a place for tools like mothur that can analyze amplicon
sequence data.

Of course this does not mean that such tools will remain static. We see
three key areas that we will continue to help the field to move forward.
First, just as we adapted through the transitions from Sanger to 454 to
MiSeq and PacBio sequencing platforms (60--62), we must learn whether
data from Oxford Nanopore can be an alternative sequencing approach that
generates sequence data that is the same quality as existing approaches;
thus far, the approach has significant shortcomings for sequencing 16S
rRNA gene sequences (74). As with the earlier platforms, we must better
understand its error profile so that sequencing errors can be corrected.
We have learned that moving forward requires that we maintain or improve
sequence quality. No doubt, datasets and read lengths will improve, but
these advances should not be made at the cost of data quality. Second,
with these improvements, we will need to continue to improve our
algorithms. We have already seen that attempts to use low quality MiSeq
and HiSeq data caused computational problems leading to the creation of
open and closed reference clustering methods, which attempted to
circumvent those problems (75, 76). Unfortunately, comparative analyses
showed that these methods fail relative to \emph{de novo} clustering
methods (47, 48). More work is needed to improve reference-based
clustering methods so that larger datasets can be analyzed without
sacrificing quality. Finally, there are ongoing controversies that need
further exploration. These include the validity and utility of amplicon
sequence variants (77), the wisdom of removing low frequency sequences
(78), and methods of identifying and removing contaminant 16S rRNA gene
sequences (79, 80). With each of these areas of development, the broader
community can count on our same data-driven approach to answer these
questions. It is common for researchers to comment that they pick a
specific method or deviate from a suggestion because they ``like how the
data look''. When pressed for an objective definition of how they know
the data look ``right'', they go quiet. Through the use of data where we
actually know what looks right and objective metrics of quality, we will
continue to base recommendations on data rather than a gut feeling.

@@

The development of mothur would not be possible without the
contributions of its many supporters, developers, and users. Although
not a complete list, I would be remiss if I did not express my gratitude
to Sarah Westcott and the other members of my research group, Jo
Handelsman, Mitch Sogin, Susan Huse, Vincent Young, Lita Proctor, Kendra
Maas, and Marcy Balunas for their unique contributions to the continued
development of mothur.

@@

\textbf{Figure 2. The mothur homepage.} From the mothur home page at
www.mothur.org, users can download mothur, access a user forum, navigate
a wiki with extensive documentation, join the mothur facebook group, and
subscribe to the mothur mailing list.

>>>>

Looking back on scientific journeys can be instructive to others who are
overwhelmed at the prospect of looking forward at their careers (1--5).
By no means is my scientific journey over, but since 2002 I have been on
a journey that I did not realize I was on. Now that the paper
introducing the mothur software package is ten years old and has become
the most cited paper published by \emph{Applied and Environmental
Microbiology} (7) (Figure 1), it is worth stepping back and using the
continued development of mothur as a story that has parallels to many
other research stories.

I fondly recall preparing a poster for the 2002 meeting of research
groups supported by the NSF-supported Microbial Observatories Program. I
wanted to triumphantly show that I had sequenced more than 600 16S rRNA
gene sequences from a single 0.5-g sample of Alaskan soil. This was
greater sequencing depth than anyone else had achieved for a single
sample. As I was preparing the poster, I walked into the office of Jo
Handelsman, my postdoctoral research advisor at the University of
Wisconsin, and laid out the outline for the poster. She asked if I could
add one of those ``curvy things'', a rarefaction curve, to show where I
was in sampling the community. Rarefaction curves and attempts to
estimate the taxonomic richness of soil had become popular because of
the impactful review by Jennifer Hughes and her colleagues (8). Their
seminal paper introduced the field to operational taxonomic units
(OTUs), rarefaction curves, and richness estimates. I do not recall
whether my poster had a rarefaction curve on it, but Jo's question and
that review article primed my career.

\textbf{\emph{Introducing DOTUR and friends.}} When Jo asked me to
generate a rarefaction curve for the poster, the request was not
trivial. How would I bin the sequences into OTUs? Hughes and her
colleagues did it manually and with fewer than 300 sequences. Although I
could possibly do that for my 600 sequences, my goal was to generate
1,000 sequences from the sample and to repeat that sampling effort with
other samples. I needed something that could be automated. Furthermore,
the software that Hughes used to build rarefaction curves, EstimateS
(4), required a series of tedious data formatting steps to perform the
analyses we were interested in performing. I had found my first problem.
How would I assign sequences to OTUs and use that data to estimate the
richness and diversity of a sample? The second problem would involve
comparing the abundance of OTUs found in one sample to another sample.
The solution to the first problem, DOTUR (Distance-based OTUs and
Richness), took us two years to develop (9). DOTUR did two things: given
a matrix quantifying the genetic distance between pairs of sequences, it
would cluster those sequences into OTUs for any distance threshold to
define the OTUs and then it would use the frequency of each OTU to
calculate a variety of alpha diversity metrics. The solutions to the
second problem would come from our work to develop software including
\(\int\)-LIBSHUFF (10), SONS (Shared OTUs and Similarity) (11), and
TreeClimber (12). Around the same time, Catherine Lozupone and Rob
Knight were developing their UniFrac tools to compare communities with a
phylogenetic rather than OTU-based approach (13, 14). With these tools,
the field of microbial ecology had a quantitative toolbox for describing
and comparing microbial communities. Along the way Jo and I would
demonstrate the utility of such tools for answering questions like how
many OTUs were there in that sample of Alaskan soil and how many
sequences were needed to sample each of those OTUs (15)? Where were we
in the global bacterial census (16)? How does the word usage of
\emph{Goodnight, Moon} compare to that of \emph{Portrait of a Lady} and
more importantly how is this relevant to microbial ecology (17)? Most
edifying were the more than 2,400 papers that used DOTUR, SONS,
TreeClimber, or \(\int\)-LIBSHUFF to facilitate their own research
questions (Web of Science, October 1, 2019). Had we waited to solve all
of the problems that plagued 16S rRNA gene sequencing, we would still be
waiting.

It is important to remember that we knew there were many problems with
16S rRNA gene sequencing. We knew there were biases from extractions and
amplification (18--23). We knew there were chimeras (24--27). We knew
that bacteria varied in their \emph{rrn} copy number. Generating a
distance matrix was a prerequisite to using my tools. This was not
trivial, but by cobbling together other tools it was possible. We would
assemble, trim and correct Sanger sequence reads using Chromas or STADEN
(28), align the sequences using ClustalW (29) or ARB (30), check for
chimeras using partial treeing or Bellerephon (27), and calculate a
pairwise distance matrix using DNADIST from the PHYLIP package (31). At
the time, we knew that we only had a loose concept of a species based on
these distances (32). We hoped that an OTU defined as a group of
sequences more than 97\% similar to each other would be a biologically
meaningful unit regardless of whether it fit our notion of a bacterial
species. At the time, I felt that the biggest problems that I could
solve were how to cluster the sequences into OTUs and how to use those
clusterings to test our hypotheses. The only tool available at the time
that automated the clustering step was FastGroup, which implemented an
approximation of the single linkage algorithm (33). The high cost of
sequencing was also an impediment to experimentation and analysis in
microbial ecology. It was rare for a study design to have experimental
replicates so that one could perform a statistical test to compare
treatment groups. For example, in our testing we frequently used a
dataset comparing Scottish soils from Alison McCaig and colleagues (34).
This dataset consisted of two experimental groups, each replicated three
times with 45 sequences per replicate. Although great focus has been
placed on the depth of sampling afforded by 454 and Illumina sequencing,
the true benefit of the modern sequencing platforms is the ability to
affordably sequence a large number of technical and biological
replicates. In my opinion, this expansion in the number of replicates
more than makes up for the potential limitations incurred by their
shorter read lenght. In spite of the many technical challenges, we had
excuses and heuristics to solve problems that served our needs. It is
telling that a recent review of ``best practices'' in generating and
analyzing 16S rRNA gene sequences shows that we still have not solved
many of these issues and that we have, of course, identified additional
problems (35).

As we developed these tools, I found a unique niche in microbiology. My
undergraduate and graduate training as a biological engineer prepared me
to think about research questions from a systems perspective, to think
quantitatively, and to understand the value of using computer programs
to help solve problems. As an undergraduate student, I learned the
Pascal programming language and promptly forgot much of it. Although it
was a good language for teaching programming concepts, it did not catch
on outside of the classroom. Later, I learned MATLAB. Because it was an
expensive commercial programming environment and never caught on with
biologists, I also forgot much of it. Even if I forgot the programming
syntax of these languages, what learning these languages taught me was
the logic and structure of programming. As a postdoc, I would use this
background to learn the Perl programming language to better understand
how LIBSHUFF (i.e.~LIBrary SHUFFle), a tool for comparing the structure
of two communities, worked since it was written in Perl (36). After
writing my own version of LIBSHUFF, \(\int\)-LIBSHUFF, and seeing the
speed of the version written in C++ by my collaborator, Bret Larget, I
converted my Perl version of DOTUR into C++. At the time, the conversion
from Perl to C++ seemed like an academic exercise to learn a new
language. My Perl version of DOTUR took a minute or so to process the
final collection of 1,000 sequences and the C++ version took seconds.
Was that really such a big difference? In hindsight, as we now process
datasets with tens of millions of sequences, the decision to learn C++
was critical. The ability to pick up computer languages to solve
problems, enabled by my prior training in engineering, was a skill that
was virtually unheard of in microbiology. Today, researchers without the
ability to program are at a significant disadvantage (37).

\textbf{\emph{Introducing mothur.}} Shortly after DOTUR was published, I
received an email from Mitch Sogin, a scientist at the Marine Biology
Laboratory (Woods Hole, MA), who asked whether DOTUR could handle more
than a million sequences. Without answering his question, I asked where
he found a million sequences. Little did I know that his email would
represent another pivot in the development of these tools and my career.
His group would be the first to use 454 sequencing technology to
generate 16S rRNA gene sequences (38). Although DOTUR could assign
millions of sequences to OTUs, it was slow and required a significant
amount of RAM. As I left my postdoc to start my independent career
across the state from Sogin's lab at the University of Massachusetts in
Amherst, my plan was to rewrite DOTUR, SONS, \(\int\)-LIBSHUFF, and
TreeClimber for the new world of massively parallelized sequencing. The
new tool would become mothur.

Milling about at a poster session at the 2007 ASM General Meeting in
Toronto, I ran into Mitch who asked what my plans were for my new lab. I
told him that I wanted to make a tool like ARB, a powerful database tool
and phylogenetics package (30), but for microbial ecology analysis. His
retort was, ``You and what army?'' Up to that point, I had written every
line of code and been answering many emails from people asking for help.
He was right, I would need an army. It would be difficult, but I needed
to learn to let go and share the development process with someone else.
My ``army'' ended up being Sarah Westcott who has worked on the mothur
project from its inception. Today, mothur is nearly 200,000 lines of
code and Sarah has touched or written nearly every line of it. Beyond
writing and testing mothur's code base, she has become a conduit for
many who are trying to learn the tools of microbial ecology. She
patiently answers questions via email and on the package's discussion
forum (\url{https://forum.mothur.org}). The community and I are lucky
that Sarah has stayed with the project for more than a decade. To be
honest, such dependency on a single person makes the project brittle. In
hindsight, it would have been better to have developed mothur with more
of an ``army'' or team so that there is overlap in people's
understanding of how mothur works. Although a distributed team approach
might work in a software engineering firm, it is not practical in most
academic environments where there is limited funding. There are
certainly projects that make this work, but they are rare.

@@

step through the project's website (\url{https://www.mothur.org}).
Third, as I mentioned above, there has been a plethora of methods

@@

\textbf{\emph{Challenges of making open source count.}} Anyone can post
code to GitHub with a permissive license and claim to be an open source
software developer. Far more challenging is engaging the target
community to make contributions to that code. Frankly, we have struggled
to expand the number of people that make contributions to the mothur
code base. One challenge we face is that if we looked to others to
contribute code to mothur, they would need to know C++. Given the
paucity of microbiologists that can program in a compiled language like
C++, expecting that community to provide contributors who can write code
in a syntax that prizes execution efficiency over developer efficiency
was not realistic. In contrast, the QIIME development team could be more
distributed because their code base was primarily written in Python,
which prizes developer efficiency over execution efficiency. QIIME is a
series of wrappers that allow users to execute other developers' code,
making the use of a scripting language like Python attractive. Their
choices resulted in many tradeoffs that have impacted ease of
installation, usability, execution speed, and flexibility. If we were
offered funding to rewrite mothur, we would likely rewrite it as an R
package that leaned heavily on the R language's C++ interface packages.
Of course, such choices are always best in hindsight. When we started
developing mothur, the ability to interface between scripting languages
like R and Python and C++ code was not as well developed as it is today.
For example, the modern version of the Rcpp package was first released
in 2009 and its popularity was not immediate (66). The development of
mothur has been a product of the environment that it was created in.
Although these decisions have largely had positive outcomes, there have
been tradeoffs that caused us to sacrifice other goals.

@@

\textbf{\emph{Failed experiments.}} If we never failed, we would not be
trying hard enough. Over the past decade we have tried a number of
experiments to improve the usability and utility of mothur. One of our
first experiments was to use mothur to generate standard vector graphic
(SVG)-formatted files of heatmaps and Venn diagrams depicting the
overlap between microbial communities. Such visuals were helpful for
exploring data; however, I quickly realized that I would never put a
mothur-generated figure into a manuscript I wrote. Such visuals require
far too much customization to be publication-quality. Although QIIME has
incorporated visualization tools through the Emperor package (67), the
challenge of users taking default values has downsides, especially when
those defaults do not follow good data visualization principles. For
example, ordinations with black backgrounds and 3-D ordinations in a 2-D
medium now litter the literature. Instead, we have encouraged users to
use R packages to visualize mothur-generated results using the minimalR
instructional materials that I have developed
(\url{http://www.riffomonas.org/minimalR/}). A second experiment was the
creation of a graphical user interface (GUI) for running mothur. Forcing
users to interact with mothur through the command line has been a
significant hurdle for many (Figure 3). Unfortunately, the development
effort required to create and maintain a GUI is significant and there is
limited funding for such efforts. The newest version of QIIME (starting
with version 2.0.0) has emphasized interaction with the tools through a
GUI (68) and the related QIITA project offers a web-based GUI (69). It
remains to be seen how this experiment will go. Another downside of
using a GUI is that there is a risk that reproducibility will suffer if
users do not have a mechanism to document their mouse clicks. A
significant downside for web interfaces is the frequent inability to
document or return to old versions of software and databases. As was
experienced with greengenes, if the website is terminated, reproducing
old analyses becomes impossible. In mothur, documentation of commands
and parameter values is explicit in users can provide a file with a list
of commands and the software returns a logfile with all commands and
output recorded. Given the heightened focus on reproducibility in recent
years, we have extended significant effort in developing instructional
materials teaching users how to organize, document, and execute
reproducible pipelines that allow a user to go from raw sequence data to
a compiled manuscript (65, 70). A final example of a failed experiment
was a collaboration with programmers through Google Summer of Code to
develop commands in mothur that ran the random forest and SVM machine
learning algorithms. Similar to the challenges of developing attractive
visuals, fitting the algorithms' hyperparameters, testing, and deploying
the resulting models require a significant amount of customization.
Furthermore, machine learning is an active area of research where
methods are still being developed and improved. Thankfully, there are
numerous R and Python packages that do a better job of developing these
models (71, 72). In each of our ``failed'' experiments, the real
problems were straying from what mothur does well and failing to grasp
what we really wanted the innovation to do.

@@

As with anything in science, methods become \text{pass\'e}. When we
first developed mothur, T-RFLP and DGGE were still commonly used. Today
it would be hard to argue that data from those methods meaningfully
advance a study relative to what one could get using 16S rRNA gene
sequence data. Looking forward, many want to claim that amplicon
sequencing is today's DGGE. They claim that researchers should instead
move on to shotgun metagenomic sequencing. It is important to note that
the two methods answer fundamentally different questions. 16S rRNA gene
sequence data describes the taxonomic composition, while metagenomic
sequence data tells a researcher about the functional potential and
genetic diversity of a community. Both tools provide important
information, but they cannot easily replace each other. Although
metagenomic data does provide highly resolved taxonomic information, its
practical limit of detection is at least an order of magnitude higher
than that of amplicon data. For example, we analyzed 10,000 16S rRNA
sequences from each of about 500 subjects (73). We can think of this as
representing about 1,000,000 genome equivalents (10,000 16S rRNA
genes/subject x 500 subjects / 5 16S rRNA gene sequences/genome).
Assuming a genome is 4 Mbp, this would represent a sequencing depth of 4
Tbp. Although such a sequencing effort is technically possible, the cost
of such an endeavor would be considerable and unlikely to be pursued by
most researchers. We estimate that generating and sequencing the
libraries at the University of Michigan sequencing core would cost
approximately \$150 per library. The parallel 16S rRNA gene sequences
data would cost approximately \$8 per library. Furthermore, analyzing
such a large dataset with an approach that captures the full genetic
diversity of the community would be financially and technically
prohibitive. Going forward, sequencing technologies will continue to
evolve to capture longer and more high quality data and there will
always be a need for characterizing the taxonomic diversity of microbial
communities. With this in mind, there will always be a place for tools
like mothur that can analyze amplicon sequence data.

Of course this does not mean that such tools will remain static. We see
three key areas that we will continue to help the field to move forward.
First, just as we adapted through the transitions from Sanger to 454 to
MiSeq and PacBio sequencing platforms (60--62), we must learn whether
data from Oxford Nanopore and other developing sequencing technologies
can be an alternative sequencing approach that generates sequence data
that is the same quality as existing approaches; thus far, the approach
has significant shortcomings for sequencing 16S rRNA gene sequences
(74). As with the earlier platforms, we must better understand its error
profile so that sequencing errors can be corrected. We have learned that
moving forward requires that we maintain or improve sequence quality. No
doubt, datasets and read lengths will improve, but these advances should
not be made at the cost of data quality. Second, with these
improvements, we will need to continue to improve our algorithms. We
have already seen that attempts to use low quality MiSeq and HiSeq data
caused computational problems leading to the creation of open and closed
reference clustering methods, which attempted to circumvent those
problems (75, 76). Unfortunately, comparative analyses showed that these
methods fail relative to \emph{de novo} clustering methods (47, 48).
More work is needed to improve reference-based clustering methods so
that larger datasets can be analyzed without sacrificing the quality of
OTU assignments. Finally, there are ongoing controversies that need
further exploration. These include the validity and utility of amplicon
sequence variants (77), the wisdom of removing low frequency sequences
(78), and methods of identifying and removing contaminant 16S rRNA gene
sequences (79, 80). With each of these areas of development, the broader
community can count on our same data-driven approach to answer these
questions. It is common for researchers to comment that they pick a
specific method or deviate from a suggestion because they ``like how the
data look''. When pressed for an objective definition of how they know
the data look ``right'', they go quiet. Through the use of data where we
actually know what looks right and objective metrics of quality, we will
continue to base recommendations on data rather than a gut feeling.

@@

The development of mothur would not be possible without the
contributions of its many supporters, developers, and users. Although
not a complete list, I would be remiss if I did not express my gratitude
to Sarah Westcott and the other members of my research group, Jo
Handelsman, Mitch Sogin, Susan Huse, Sarah Highlander, Vincent Young,
Lita Proctor, Kendra Maas, and Marcy Balunas for their unique
contributions to the continued development of mothur.

@@

\textbf{Figure 2. The mothur homepage.} From the mothur home page at
www.mothur.org, users can download mothur, access a user forum, navigate
a wiki with extensive documentation, find blog posts that provide
additional examples of how to use mothur, join the mothur facebook
group, and subscribe to the mothur mailing list.
